---
title: "Open Plaques - UK"
author: "Esraa Dandash"
date: "01/03/2023"
output: github_document

references:
- id: openplaques2023
  title: openplaques.org database
  author:
  - literal: openplaques.org
  URL: 'https://openplaques.org/data/'
  type: webpage
  issued:
    year: 2023

- id: duke-williams2023
  title: INST0065 course notes
  author:
    - family: Duke-Williams
      given: Oliver
  URL: 'https://moodle.ucl.ac.uk/course/view.php?id=27657'
  issued:
    year: 2023

- id: r-error-messages2023
  title: r-error-messages
  author:
  - literal: programmingr.com
  URL: 'https://www.programmingr.com/r-error-messages/error-in-contrib-urlrepos-source-r-markdown/#'
  type: webpage
  issued:
    year: 2023

- id: kaggle2023
  title: Exploring Blue Plaques Dataset
  author:
  - family: ResidentMario
  URL: 'https://www.kaggle.com/code/residentmario/exploring-blue-plaques'
  type: webpage
  issued:
    year: 2023

- id: questionpro2023
  title: Demographic Examples
  author:
  - literal: QuestionPro
  URL: 'https://www.questionpro.com/blog/demographic-examples/'
  type: webpage
  issued:
    year: 2023

- id: rjournal2013
  title: RJournal - Spatial Visualization with ggplot2
  author: Kahle
  URL: 'https://journal.r-project.org/archive/2013-1/kahle-wickham.pdf'
  type: article-journal
  issued:
    year: 2013

- id: bookdown2023a
  title:  Geocode and Map the British Library's Newspaper Collection
  author: Yann Ryan
  URL: 'https://bookdown.org/yann_ryan/r-for-newspaper-data/mapping-with-r-geocode-and-map-the-british-librarys-newspaper-collection.html'
  type: webpage
  issued:
    year: 2023

- id: bookdown2023b
  title: Regular Expressions
  author: Peng
  URL: 'https://bookdown.org/rdpeng/rprogdatascience/regular-expressions.html'
  type: webpage
  issued:
    year: 2023
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Data

The data for this markdown file are taken from [https://openplaques.org/data/](openplaques.org) (@openplaques2023)

see also (@r-error-messages2023)

```{r load-data, include=FALSE}
# CHANGE THE WD HERE TO INCLUDE THE ROOT DIRECTORY OF YOUR DATA
setwd("/Users")
plaques <- read_csv("open-plaques-gb-2021-05-14.csv")
```

For initial testing, we can look at a sample of data, to confirm that it has loaded properly

```{r test-data, include=TRUE}
# set to include=TRUE for testing, and then revert to include=FALSE
head(plaques)
```


## Main Objective

In this file we aim to analyse different aspects in the plaques dataset using a number of R libraries such as ggplot, ggmap and dplyr. We wrangle and aggregate the data, visualise it, identify patterns in it, analyse room for growth in the data and explore a number of particularly helpful references that can be of value to anyone looking to produce similar R functions and visualisations.

## Data Sources

The specific data sources for this dataset were not explicitly included on the website above. The data could have been sourced from various places, including public records, historical archives, online databases, or crowdsourced efforts. (@kaggle2023)

## Data Overview

The dataset at hand comprises information about plaques, specifically focusing on their attributes, inscriptions, and associated subjects. It provides a comprehensive collection of data related to these commemorative markers. The dataset serves as a valuable resource for exploring and studying historical plaques and the individuals they commemorate.
 The full list of data types in the dataset is provided in the table below:
 
| Field Name                | Type     | Meaning                                       |
|---------------------------|----------|-----------------------------------------------|
| id                        | Numeric  | Unique identifier for each plaque              |
| machine_tag               | Character| Machine-readable tag associated with the plaque|
| title                     | Character| Title of the plaque                           |
| inscription               | Character| Inscription on the plaque                      |
| latitude                  | Numeric  | Latitude coordinate of the plaque              |
| longitude                 | Numeric  | Longitude coordinate of the plaque             |
| as_wkt                    | Character| Well-Known Text representation of the plaque's location |
| country                   | Character| Country where the plaque is located            |
| area                      | Character| Area within the country where the plaque is located |
| address                   | Character| Address of the plaque                          |
| erected                   | Numeric  | Year of erection of the plaque                 |
| main_photo                | Character| URL or file path of the main photo of the plaque|
| colour                    | Character| Colour of the plaque                            |
| organisations             | Character| Organizations associated with the plaque       |
| language                  | Character| Language of the inscription on the plaque       |
| series                    | Character| Series to which the plaque belongs             |
| series_ref                | Character| Reference number of the series                 |
| geolocated?               | Boolean  | Indicates if the plaque is geolocated          |
| photographed?             | Boolean  | Indicates if the plaque is photographed        |
| number_of_subjects        | Numeric  | Total number of subjects depicted on the plaque|
| number_of_male_subjects   | Numeric  | Number of male subjects depicted on the plaque |
| number_of_female_subjects | Numeric  | Number of female subjects depicted on the plaque |
| number_of_inanimate_subjects | Numeric | Number of inanimate subjects depicted on the plaque |
| lead_subject_id           | Numeric  | Unique identifier for the lead subject          |
| lead_subject_machine_tag  | Character| Machine-readable tag for the lead subject       |
| lead_subject_name         | Character| Name of the lead subject                        |
| lead_subject_surname      | Character| Surname of the lead subject                     |
| lead_subject_sex          | Character| Sex of the lead subject                         |
| lead_subject_born_in      | Numeric  | Year of birth of the lead subject               |
| lead_subject_died_in      | Numeric  | Year of death of the lead subject               |
| lead_subject_type         | Character| Type of the lead subject                        |
| lead_subject_roles        | Character| Roles of the lead subject                       |
| lead_subject_primary_role | Character| Primary role of the lead subject                |
| lead_subject_wikipedia    | Character| Wikipedia page of the lead subject              |
| lead_subject_dbpedia      | Character| DBpedia page of the lead subject                |
| lead_subject_image        | Character| URL or file path of the image of the lead subject |
| subjects                  | Character| Subjects depicted on the plaque                 |

## Data Cleaning, Wrangling 

Data cleaning and wrangling involve preparing and transforming raw data into a clean and structured format for analysis. Some of the most commonly used ones that we leveraged in our analysis and producing the 7 plots in this file are as follows:

1. Handling missing values: Identify and handle missing data using functions like `is.na()` to retrieve and change (mutate) dataset values such as plot 2,4 and 6. 

2. Data transformation: Apply mathematical operations or create new variables using functions like `mutate()` in plots 2 and 4.

4. Filtering and subsetting: Subset data based on specific conditions using functions like `subset()` or the `filter()` function from the `dplyr` package. Select specific columns using the `$` operator or functions like `select()`. These have been all used in the 7 plots for processing the data before anlysing it.

5. Aggregating and summarizing data: Aggregating data using functions like `aggregate()` or `summarize()` from the `dplyr` package. Calculate summary statistics, group data by specific variables, or compute aggregate values as in plot 4 and plot 6.

6. Data joining: Combine datasets vertically using functions like `rbind()` as in plot 7.


## Plot 1: Analysis of Primary Roles
-----
__Objective and Methodology__: We study the top 10 most frequent "primary roles" by using the table function and sorting the frequencies in descending order. We create a subset of the data that includes only the plaques with these top 10 roles. Finally, we create a stacked bar chart using ggplot, where the x-axis represents the "primary roles", the fill color represents the "sex", and the height of the bars represents the frequency of each role. The resulting chart provides insights into the distribution of primary roles by sex.  
__Results__: A significant under-respresentation of women can be observed.
```{r plt1}
# Filter out objects with "object" sex and empty primary roles
filtered_plaques <- subset(plaques, lead_subject_sex != "object" & lead_subject_primary_role != "")

# Get the count of unique primary roles in the filtered data
unique_roles <- unique(filtered_plaques$lead_subject_primary_role)
num_roles <- length(unique_roles)

# Print the unique roles and their count (for debugging)
#print(unique_roles)
#print(num_roles)
# Get the top 10 primary roles by frequency
top_roles <- names(sort(table(filtered_plaques$lead_subject_primary_role), decreasing = TRUE)[1:10])

# Create a subset of the filtered data with only the top 10 roles
subset_plaques <- subset(filtered_plaques, lead_subject_primary_role %in% top_roles)

# Create a stacked bar chart of the top 10 roles
ggplot(subset_plaques, aes(x = factor(lead_subject_primary_role), fill = factor(lead_subject_sex))) + 
  geom_bar(position = "stack") +
  scale_fill_manual(values = c("#FFC0CB", "#89CFF0")) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1), plot.width = NULL) +
  labs(title = "Frequency of Top 10 Lead Subject Primary Roles by Sex",
       x = "Primary Role", y = "Frequency")
```

## Plot 2: Analysis of Location - By Cities (Spatial Scatterplot)
__Objective and Methodology__: Here we are using the ggmap and maps libraries to create a world map visualization with plotted points representing plaque locations. We are limiting the map to UK coordinates only to understand more about the distributions. (@rjournal2013)  
__Results__: Concentration of plaques in urban areas, particularly in major cities like London, Manchester, and Edinburgh.
```{r plt2}
library(ggmap)
library(maps)
#extract the latitude and longitude coordinates from the plaques dataset
lat <- plaques$latitude
lng <- plaques$longitude
#load the world map data 
worldmap = map_data('world')
geocorrected = plaques %>%
  mutate(lat = as.numeric(lat)) %>%
  mutate(lng = as.numeric(lng)) %>%
  filter(!is.na(lat)) %>% filter(!is.na(lng))
ggplot() +
  geom_polygon(data = worldmap,
               aes(x = long, y = lat, group = group),
               fill = 'gray90', color = 'black') +
  coord_fixed(ratio = 1.3, xlim = c(-10,3), ylim = c(50, 59)) +
  theme_void() +
  geom_point(data = geocorrected,
             aes(x = as.numeric(lng),
                 y = as.numeric(lat)),
                 alpha = .7) +
  scale_size_area(max_size = 8) +
  scale_color_viridis_c() +
  theme(legend.position = 'none') +
  theme(title = element_text(size = 12))
```


## Plot 3: Analysis of Inscription Length
__Objective and Methodology__: Here, we are calculating the length of each inscription in the plaques dataset using the nchar function and storing it in the insc_length variable. We then determine the range of inscription lengths using the range function and print the resulting limits.  
__Results__: Vast majority of inscriptions are less than 800 characters.
```{r plt3}
insc_length <- nchar(plaques$inscription)
x_limits <- range(insc_length)
print(x_limits)
ggplot(data = plaques, aes(x = str_length(inscription))) +
  geom_histogram(binwidth = 20, color = "blue", fill = "lightblue") +
  xlim(x_limits) +
  labs(title = "Length of Inscription",
       x = "Length of Inscription (characters)",
       y = "Count")
```

## Plot 4: Analysis of Documentation Techniques
__Objective and Methodology__: Here, we assess the documentation quality of the plaques in the dataset. The code calculates the documentation level based on the presence or absence of information in the lead_subject_wikipedia, lead_subject_dbpedia, photographed, and geolocated fields. It then groups the plaques by their documentation level and creates a bar chart to visualize the count of plaques in each documentation level category. This analysis helps evaluate the completeness and reliability of the documentation associated with the plaques, providing insights into the overall quality of the dataset's information.  
__Results__: Vast majority of inscriptions are relatively good in terms of documentation. Further data investigation needs to be done to udnerstand why it is showing most as unknown.
```{r plt4}
library(dplyr)
plaques <- rename(plaques, `photographed` = `photographed?`)
plaques <- rename(plaques, `geolocated` = `geolocated?`)

plaques_summary <- plaques %>%
  summarize(
    lead_subject_wikipedia_empty = sum(is.na(lead_subject_wikipedia)),
    lead_subject_wikipedia_not_empty = sum(!is.na(lead_subject_wikipedia)),
    lead_subject_dbpedia_empty = sum(is.na(lead_subject_dbpedia)),
    lead_subject_dbpedia_not_empty = sum(!is.na(lead_subject_dbpedia)),
    photographed_true = sum(photographed),
    photographed_false = sum(!photographed),
    geolocated_true = sum(geolocated),
    geolocated_false = sum(!geolocated),
  )
glimpse(plaques_summary)

plaques %>%
  mutate(documentation_level = case_when(
           !is.na(lead_subject_wikipedia) + !is.na(lead_subject_dbpedia) + photographed + geolocated == 0 ~ "poor",
           !is.na(lead_subject_wikipedia) + !is.na(lead_subject_dbpedia) + photographed + geolocated == 1 ~ "fair",
           !is.na(lead_subject_wikipedia) + !is.na(lead_subject_dbpedia) + photographed + geolocated == 2 ~ "good",
           !is.na(lead_subject_wikipedia) + !is.na(lead_subject_dbpedia) + photographed + geolocated == 3 ~ "very good",
           !is.na(lead_subject_wikipedia) + !is.na(lead_subject_dbpedia) + photographed + geolocated == 4 ~ "excellent",
           TRUE ~ "unknown" )) %>%

  group_by(documentation_level) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = documentation_level, y = count, fill = documentation_level)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3", "#ff7f00")) +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1), plot.width = NULL) +
  labs(x = "Documentation Level", y = "Count", fill = "Documentation Level Category")
```

## Plot 5: Analysis of Colour
__Objective and Methodology__: Here, we  analyze the frequency count of plaque colours in the dataset and create a bar chart to visualize the top 10 colours.The code generates a bar chart using ggplot2 with the x-axis representing the reordered colour categories based on their count, the y-axis representing the count, and the fill colour matching the value of each colour category. The chart is displayed with a flipped coordinate system, and the title, x-label, and y-label are added for better visualization and understanding.  
__Results__: The visualizations reveal a diverse range of plaque colors, indicating variations in materials and styles used for plaques. The most sued colour however is blue.
```{r plt5}
# Generate a frequency count for the "colour" field and select top 10 colours
colour_counts <- data.frame(table(plaques$colour))
colnames(colour_counts) <- c("Colour", "Count") 
# print(colour_counts)
colour_counts <- head(colour_counts[order(colour_counts$Count, decreasing = TRUE),], 10)
# slate is an unknown colour so I will replace it, not wowking still 
# Create the bar chart with fill color matching the value of Colour, mapping still not working 
ggplot(colour_counts, aes(x=reorder(Colour, Count), y=Count, fill=Colour)) +
  geom_bar(stat="identity") +
  coord_flip() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1), plot.width = NULL) +
  ggtitle("Frequency Count of Top 10 Plaque Colours") +
  xlab("Colour") +
  ylab("Count")
```

## Plot 6: Analysis of year of installation/erection
__Objective and Methodology__: Here, we  analyze the count of plaques by year of erection and lead subject sex, and create a visualization after trying to understand the data using `summarize`. The resulting data is used to create a scatter plot and line plot using ggplot2. The x-axis represents the year of erection, the y-axis represents the count, and the color of the points and lines is determined by the lead subject sex. (@questionpro2023)

The visualization is further enhanced by using facet_wrap() to create separate panels for each lead subject sex, and scale_x_continuous() to limit the x-axis range from 1800 to 2023 for better visibility and interpretation.  
__Results__: Most of the plaques are quite recent regardless of their types (object, male or female).
```{r plt6}
library(dplyr)

plaques %>%
  filter(!is.na(erected)) %>% 
  filter(!is.na(lead_subject_sex)) %>% 
  group_by(lead_subject_sex, erected) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = erected, y= count, color = lead_subject_sex)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ lead_subject_sex, ncol = 1) +
  scale_x_continuous(limits = c(1800, 2023)) +
  labs(x = "Year of erection", y = "Count") 
```
## Plot 7: Analysis of Specific Role Distributions
__Objective and Methodology__: The purpose of this plot is to analyze the distribution of artist plaques across different areas and create a bar chart showing the counts of artist plaques in the top 10 and bottom 10 areas. The code selects rows from the dataset where the 'lead_subject_primary_role' column contains the word "artist" and stores them in a separate data frame. It then calculates the counts of artist plaques for each area using regular expression (@bookdown2023b) and lead subject primary role combination. The top and bottom areas are determined based on these counts, and a new data frame is created combining the top and bottom areas. Finally, a bar chart is generated using ggplot2, displaying the counts of artist plaques for each area, with the areas reordered by count and bars colored in sky blue. The chart also includes a title, labels for the axes, and a dashed line representing the median count.  

__Results__: London has the most number of plaques followed by Middleton with a big difference, which makes since since there are many art galleries and a keen apporach to preserve arts in the capital. Tullylish ahs the smallest number of Plaques for artists.
```{r plt7}
library(dplyr)

# Select only the rows where the 'lead_subject_primary_role' column contains the word "artist"
artist_plaques <- plaques[grep("artist", plaques$lead_subject_primary_role, ignore.case = TRUE), ]
artist_areas <- artist_plaques %>%
  select(area, lead_subject_primary_role)

# Create a  table of the counts of each area and lead subject primary role combination
area_counts <- table(artist_areas)

# Get the total counts for each area
area_totals <- rowSums(area_counts)

# Create a data frame of the areas and their corresponding counts
area_df <- data.frame(area = names(area_totals), count = area_totals)

# Sort the data frame by the counts in descending order
area_df <- area_df %>%
  arrange(desc(count))

# Select the top 10 and bottom 10 areas
top_areas <- head(area_df, 10)
bottom_areas <- tail(area_df, 10)

# Combine the top 10 and bottom 10 data frames
plot_df <- rbind(top_areas, bottom_areas)

# Create a bar chart of the counts for each area
ggplot(plot_df, aes(x = reorder(area, count), y = count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  ggtitle("Top 10 and Bottom 10 Areas with Artist Plaques") +
  xlab("Area") +
  ylab("Count") +
  coord_flip() +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom") +
  geom_hline(yintercept = median(plot_df$count), linetype = "dashed", color = "red")
```


## Summary of Notable Features and Patterns in Visualisations
Through the analysis and visualizations of the plaqes dataset, we have observed notable patterns and insights:  
* The dataset shows a notable concentration of plaques in urban areas, particularly in major cities like London, Manchester, and Edinburgh. This raise the questions of whether or not that is due to the increase of interested organisations in theses areas or the desire to show them to more people.  
* The visualizations reveal a diverse range of plaque colors, indicating variations in materials and styles used for plaques.  
* Another interesting pattern is the presence of multiple languages on the plaques, reflecting the multicultural heritage and history of the regions.  
* The dataset also highlights the significance of gender representation, with a number of plaques dedicated to notable women and their contributions.  


In the data itself, a significant pattern observed is the high number of plaques associated with historical landmarks, such as castles, cathedrals, and famous buildings although it has not been visualised in this R project. 
## Further Development 
* Demographically, the datset could be more developed by including ethinicities, educational backgrounds and other parameters that can help identify representation of all parties.  
* Improving documentation by including correct, full information of the plaques and checking for null values.   
* Leveraging natural language processing techniques could help gain insights into the inscriptions, unraveling hidden meanings and historical contexts.    
* The `erected` variable in particular is missing most of its data, and I believe relating the ages of the plaques to historical events and timelines could provide valuable historical context and facilitate a better conclusion on whether organisations are producing ones in an increasing or declining manner.  


# Bibliography
